---
title: "Bacteremia study"
output:
  html_document:
    df_print: paged
---

This document describes the bacteremia data set for use as an example in the IDA project. 

# 1. Introduction

Ratzinger et al (2014) performed a diagnostic study in which age, sex and 49 laboratory variables can be used to diagnose bacteremia status  of a blood sample using a multivariable model.  Between January 2006 and December 2010, patients with the clinical suspicion to suffer from bacteraemia were included if blood culture analysis was requested by the responsible physician and blood was sampled for assessment of haematology and biochemistry. The data consists of 14,691 observations from different patients.

Our version of this data was slightly modified compared to original version, and this modified version was cleared by the Medical University of Vienna for public use (DC 2019-0054).

In the original paper describing the study ("Ratzinger-2013-PlosOne.PDF"), a machine learning approach was taken to diagnose a positive status of blood culture. The true status was determined for all blood samples by blood culture analysis, which is the gold standard but . Here we will make use of a multivariable logistic regression model.


## Load the data

```{r}
bact<-read.csv("../data/Bacteremia_public.csv")
dim(bact)
```

## Intended statistical analysis

Suppose we are interested to fit a multivariable logistic regression to predict positive blood culture status based on the 49 laboratory variables, age and gender.

The 49 laboratory variables  are listed along with units of measurement in the original paper. `ID` is a (permuted) patient identifier, and laboratory variables are the variables from `MCV` to `WBC`. `BloodCulture` is the result of blood culture analysis and coded as string in `no` and `yes`.

```{r}
colnames(bact)
```

A data dictionary is available:

```{r}
bact_dd<-read.csv2("bacteremia-DataDictionary.csv")
View(bact_dd)
```



Briefly, initial data analysis will focus on evaluating the distribution of variables and possible transformations suggested to reduce disproportional impact of outliers, reduce the number of variables considered for analysis by evaluating collinearity, and consider exclusion of variables based on excessive missing values.

For diagnosis a parsimonious model is desired. The optimal analysis strategy (preselection of predictors, data-driven selection algorithm) cannot easily be determined without proper initial data analysis. 

