---
title: "Pre-process the CRS data set"
output:
  md_document:
    variant: markdown_github
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# About

This notebook is for final processing of data prior to IDA e.g. completing missing meta-data, labels, etc. 

# Set up

Set up packages and path to the data set. 

```{r setupR}
library(rmarkdown)
library(ggplot2)
library(dplyr)
library(tidyr)
library(Hmisc)
library(here)

```


# Read data

TODO:add variable to indicate measurement type:

* med history 
* outcome
* Add additional metadata on measurement schedule

Load the CRS dataset.

```{r}
load(here("data","crs.Rdata"))
```

# Check data

Check data contents and complete any misssing information including labels or units. 

display the data set contents. 

```{r}
Hmisc::contents(crs) 
```

Add missing label for BMI. 

```{r}
a_crs <- crs %>%
  Hmisc::upData(labels = c(bmi = 'body mass index'))
```

# Save data sets

Save the updated data set. 

```{r}
save(crs, a_crs, file = here("data","a_crs.Rdata"))
```

# Session info

```{r}
sessionInfo()
```


