# Univariate distribution checks

This section reports a series of univariate summary checks of the bacteremia dataset.

```{r, message =FALSE, warning =FALSE , echo=FALSE}
library(here)
library(tidyverse)
library(Hmisc)
source(here("R", "ida_plot_univar.R"))  ## function to plot univariate summaries. 
source(here("R", "ida_plot_univar_orig_vs_trans.R"))  ## function for side-by-side comparison of original vs. transformed scale, calls ida_plot_univar.R
source(here("R", "ida_trans.R"))  ## function to determine transformation (log(x+c) or identity). 

## Load the dataset.
load(here::here("data", "bact_env_b.rda"))

```

## Data set overview

Using the [Hmisc](https://cran.r-project.org/web/packages/Hmisc/) describe function, we provide an overview of the data set. The descriptive report also provides histograms of continuous variables. For ease of scanning the information, we group the report by measurement type. 

### Demographic variables

```{r desc_b_bact_1, message = FALSE, results='asis', warning=FALSE, , echo=FALSE}
b_bact %>%
  dplyr::select(all_of(demog_vars)) %>%
  Hmisc::describe(descript = "Demographic variables") %>%
  Hmisc::html(size = 80)
```


### Structural covariates and key predictors

```{r desc_b_bact_2, message = FALSE, results='asis', warning=FALSE, , echo=FALSE}
b_bact %>%
  dplyr::select(all_of(c(structural_vars, key_predictors))) %>%
  Hmisc::describe(descript = "Structural covariates and key predictors") %>%
  Hmisc::html(size = 80)
```

### Further variables related to leukocyte types and leukocyte ratios

```{r desc_b_bact_3, message = FALSE, results='asis', warning=FALSE, echo=FALSE}
b_bact %>%
  dplyr::select(all_of(c(leuko_related_vars,leuko_ratio_vars))) %>%
  Hmisc::describe(descript = "Leukocyte related variables and leukocyte ratios") %>%
  Hmisc::html(size = 80)
```

### Kidney function related variables

```{r desc_b_bact_4, message = FALSE, results='asis', warning=FALSE, echo=FALSE}
b_bact %>%
  dplyr::select(all_of(c(kidney_related_vars))) %>%
  Hmisc::describe(descript = "Kidney function related variables") %>%
  Hmisc::html(size = 80)
```

### Acute phase reaction related variables

```{r desc_b_bact_5, message = FALSE, results='asis', warning=FALSE, echo=FALSE}
b_bact %>%
  dplyr::select(all_of(c(acute_related_vars))) %>%
  Hmisc::describe(descript = "Acute phase related variables") %>%
  Hmisc::html(size = 80)
```

### Remaining variables

```{r desc_b_bact_6, message = FALSE, results='asis', warning=FALSE, echo=FALSE}
b_bact %>%
  dplyr::select(all_of(c(remaining_vars))) %>%
  Hmisc::describe(descript = "Remaining variables") %>%
  Hmisc::html(size = 80)
```

## Categorical variables

We now provide a closer visual examination of the categorical predictors. 

```{r catplot, message=FALSE, warning =FALSE , echo=FALSE}
b_bact %>%
  dplyr::select(SEX, BC) %>%
  dplyr::mutate_all(forcats::as_factor) %>%   
  dplyr::mutate_all(forcats::fct_explicit_na, "NA") %>%
  tidyr::pivot_longer(
    dplyr::everything(),
    names_to = "var",
    values_to = "value",
    values_drop_na = FALSE
  ) %>%
  dplyr::group_by(var, value) %>%
  dplyr::summarize(N = n()) %>%
  dplyr::mutate(
    freq = N / sum(N),
    pct = round((freq * 100), 1),
    axis_lab = paste0(value, ' ', '(N = ', N, ')'),
    var_label = case_when(var == "SEX" ~ "Sex",
                          var == "BC" ~ "Bacteremia")
  ) %>%
  ggplot(aes(
    x = reorder(axis_lab, pct),
    y = pct,
    label = pct
  )) +
  geom_text(nudge_y = 7) +
  geom_pointrange(aes(ymin = 0, ymax = pct), alpha = 1, size = 1, color = "grey") +
  geom_point(color = "firebrick2",
             alpha = 0.6,
             size = 3) +
  ylab("Percentage (%)") +
  scale_y_continuous(limits = c(0, 100)) +
  labs(caption = "Number of subjects with a non-missing value reported in brackets.\nNA = missing") +
  facet_wrap(~ var_label, ncol = 1, scales = "free_y") +
  coord_flip() +
  theme_minimal(base_size = 12) +
  theme(axis.title.y = element_blank(),
        panel.grid.minor = element_blank())

```


## Continuous variables 



### Suggested transformations

Next we investigate whether a transformation of continuous variables may improve any further analyses to reduce disproportional impact of highly influential points, also in multivariate summaries. We employ a function `ida_trans` for this purpose, which optimises the parameter `sigma` of the pseudo-logarithm for that purpose. The optimization targets the best possible linear correlation of the transformed values with normal deviates. If no better transformation can be found, no transformation is suggested.

```{r}
variables<- c("AGE", structural_vars, key_predictors, leuko_related_vars, leuko_ratio_vars, kidney_related_vars, acute_related_vars, remaining_vars)
unique.variables <- unique(variables)

res<-sapply(unique.variables, function(X) ida_trans(b_bact[,X])$const) #takes long, calculate once, and save?

res
```

Register transformed variables in the data set:

```{r}
for(j in 1:length(unique.variables)){
  if(!is.na(res[j])){
    newname <- paste("t_",unique.variables[j],sep="")
    newlabel <- paste("pseudo-log of",label(b_bact)[unique.variables[j]])
    names(newlabel)<-newname
    x<-pseudo_log(b_bact[[unique.variables[j]]], sigma=res[j], base=10)
    label(x)<-newlabel
    b_bact[[newname]] <- x
    upData(b_bact, labels=newlabel)
  }
}

sigma_values <- res


c_bact <- b_bact

# update variable lists - generate a second list with transformed variables replacing the originals

bact_transformed <- bact_variables

for(j in 1:length(bact_variables)){
  for(jj in 1:length(bact_variables[[j]])){
      if(!is.na(res[bact_variables[[j]][jj]])) bact_transformed[[j]][jj] <- paste("t_", bact_variables[[j]][jj], sep="")
  }
}


```


### Univariate distribution with variables using the original variable and the suggested transformations

```{r}
for(j in 1:length(unique.variables)){
  print(ida_plot_univar(b_bact, unique.variables[j], sigma=res[j], n_bars=100))
#  if(!is.na(res[j])){
#    print(ida_plot_univar(b_bact, paste("t_",variables[j],sep="")))
#  }
}
```
### Comparison of univariate distributions with and without pseudo-log transformation

The comparison is only shown for variables where a transformation is suggested.

```{r, fig.width=12}
for(j in 1:length(unique.variables)){
#  print(ida_plot_univar_orig_vs_trans(b_bact, unique.variables[j], sigma=res[j], n_bars=100))
 if(!is.na(res[j])){
   print(ida_plot_univar_orig_vs_trans(b_bact, unique.variables[j], sigma=res[j], n_bars=100))
 }
}
```

```{r}
save(list=c("c_bact", "bact_variables", "sigma_values", "bact_transformed"), 
     file=here::here("data", "bact_env_c.rda"))
```

### Univariate distribution with variables using only the original variable without the suggested transformations

```{r}
for(j in 1:length(unique.variables)){
  print(ida_plot_univar(b_bact, unique.variables[j], sigma=res[j], n_bars=100, transform = FALSE))
#  if(!is.na(res[j])){
#    print(ida_plot_univar(b_bact, paste("t_",variables[j],sep="")))
#  }
}
```

### Comparison of univariate distributions with and without pseudo-log transformation

The comparison is only shown for variables where a transformation is suggested.

```{r, fig.width=12}
for(j in 1:length(unique.variables)){
#  print(ida_plot_univar_orig_vs_trans(b_bact, unique.variables[j], sigma=res[j], n_bars=100))
 if(!is.na(res[j])){
   print(ida_plot_univar_orig_vs_trans(b_bact, unique.variables[j], sigma=res[j], n_bars=100))
 }
}
```


## Section session info

```{r, warning=FALSE, message=FALSE, echo=FALSE}
sessionInfo()
```


