# CRASH-2 {-}


```{r, echo=FALSE, warning=FALSE, message=FALSE, echo=FALSE}
## Load libraries for this chapter
library(here)
library(tidyverse)
library(Hmisc)
```

# Introduction to CRASH-2

Since a key principle of IDA is not to touch the research questions, before IDA commences the research aim and statistical analysis plan need to be in place. 
IDA may lead to an update or refinement of the analysis plan. To demonstrate the workflow and content of IDA, we created a hypothetical research aim and corresponding statistical analysis plan, which is described in more detail in the section [Crash2_SAP.Rmd](Crash2_SAP.Rmd).

**Hypothetical research aim for IDA** is to develop a multivariable model for early death (death within 28 days from injury) using nine independent variables of mixed type (continuous, categorical, semicontinuous) with the primary aim of prediction and a secondary aim of describing the association of each variable with the outcome. 

A prediction model was developed and validated based on this data set in "Predicting early death in patients with traumatic bleeding"  [Perel et al, BMJ 2012](https://doi.org/10.1136/bmj.e5166), [supplement available at]. The assumed research aim is in line with the prediction model


## CRASH-2 Description

Clinical Randomisation of an Antifibrinolyticin Significant Haemorrhage(CRASH-2)  was a large randomised placebo controlled trial among trauma patients with, or at risk of, significant haemorrhage, of the effects of antifibrinolytic treatment on death and transfusion requirement. The study is described at [the original trial website](http://crash2.lshtm.ac.uk/). A public version of the data set is found at a [repository of public data sets](http://biostat.mc.vanderbilt.edu/wiki/Main/DataSets) hosted by the Vanderbilt University's Department of Biostatistics (Prof. Frank Harrell Jr.).

The data set includes 20,207 patients and 44 variables. 

**Note:** In contrast to the analysis described in Perel et al, variables describing the economic region and the treatment allocation are missing in the public version of the data set, and while the data set contains 20,207 patients, the research paper mentions 20,127 patients having been included in the study.


## Crash2 dataset contents

### Source dataset 

We refer to the source data set as the dataset available online [here](http://biostat.mc.vanderbilt.edu/wiki/Main/DataSets)

Display the source dataset contents. This dataset is in the **data-raw** folder of the project directory. 

```{r contents_crash2,  warning=FALSE, message=FALSE, echo=FALSE, results='asis'} 
## Load the dataset from the data-raw folder. 
## Alternatively we could load directly from the url 
## http://biostat.mc.vanderbilt.edu/wiki/pub/Main/DataSets/crash2.rda
load(here::here("data-raw", "crash2.rda"))

Hmisc::html(Hmisc::contents(crash2),
            maxlevels = 10,
            levelType = 'table')

```


### Updated analysis dataset

Additional meta-data is added to the original *source* data set. We write this new modified data set back to the **data** folder after adding additional meta-data for the following variables:

* **age** - add label "Age" and unit "years". 
* **injury time** - add unit "hours". 
* **total Glasgow coma score** - add unit "points". 


At the stage we select the variables of interest to take in to the IDA phase by dropping variables we do not check in IDA.

As a cross check we display the contents again to ensure the additional data is added, and then write back the changes to the data folder in the file "data/a_crash2.rds". 

```{r contents_acrash2, warning=FALSE, message=FALSE, echo=FALSE, results='asis'} 

## Complete metadata by adding missing labels. 
## Generate adervived dataset stored in data as we are adding to the oirginal source dataset obtained. 
## select candidate predictor variables. -- See SAP

crash2_subset <-
  crash2 %>%
  dplyr::select(
    entryid,        # patient identifer
    trandomised,    # date of randomisation 
    ddeath,         # date of death
    age,            # Age (`age`, years)
    sex,            # Sex (`sex`, male or female)
    sbp,            # Systolic blood pressure (`sbp`, mmHg)
    hr,             # Heart rate (`hr`, 1/min)
    rr,             # Respiratory rate (`rr`, 1/min)
    gcs,            # Glasgow coma score (`gcs`, points)
    cc,             # Central capillary refill time (`cc`, seconds)
    injurytime,     # Time since injury (`injurytime`, hours)
    injurytype      # Type of injury (`injurytype`, 'blunt', 'penetrating' or 'blunt and penetrating')
  )


## Complete metadata by adding missing labels.
a_crash2 <- Hmisc::upData(
  crash2_subset,
  labels = c(
    age = 'Age',
    sex = "Sex", 
    injurytype = "Injury type", 
    time2death = "Time from randomization to day of death"),
  units = c(
    age = "years",
    injurytime = "hours",
    gcs = "points", 
    time2death = "days"
  )
)

## Derive outcome variable
a_crash2$time2death <-
  as.numeric(as.Date(a_crash2$ddeath) - as.Date(a_crash2$trandomised))

a_crash2$earlydeath[!is.na(a_crash2$time2death)] <-
  (a_crash2$time2death[!is.na(a_crash2$time2death)] <= 28) + 0

# +0 to transform it from TRUE/FALSE to 1/0
# NA in time2death means alive at day 28
a_crash2$earlydeath[is.na(a_crash2$time2death)] <- 0    

## Add meta data
a_crash2 <- Hmisc::upData(a_crash2 ,
               labels = c(earlydeath = 'Death within 28 days from injury'))


## Display contents
Hmisc::html(Hmisc::contents(a_crash2),
            maxlevels = 10,
            levelType = 'table')

## Write to data folder
save(a_crash2, file = here::here("data", "a_crash2.rda"))
```

## Chapter session info

```{r, warning=FALSE, message=FALSE, echo=FALSE}
sessionInfo()
```



