# Univariate distribution checks

This section reports a series of univariate summary checks of the CRASH-2 dataset.

```{r, message =FALSE, warning =FALSE , echo=FALSE}
library(here)
library(tidyverse)
library(Hmisc)
source(here("R", "ida_plot_univar.R"))  ## function to plot univariate summaries. 

# Define sets of predictors by type - helpful for selection
demog_vars <- c("age", "sex")
pyhiso_vars <- c("sbp", "hr", "rr", "gcs", "cc")
injury_vars <- c("injurytime", "injurytype")

## Load the dataset.
load(here::here("data", "a_crash2.rda"))
```

## Data set overview

Using the [Hmisc](https://cran.r-project.org/web/packages/Hmisc/) describe function, we provide an overview of the data set. The descriptive report also provides histograms of continuous variables. For ease of scanning the information, we group the report by measurement type. 

### Demographic variables

```{r desc_a_crash2a, message = FALSE, results='asis', warning=FALSE, , echo=FALSE}
a_crash2 %>%
  dplyr::select(all_of(demog_vars)) %>%
  Hmisc::describe(descript = "Demographic variables") %>%
  Hmisc::html(size = 80)
```


### Physiological measurements

```{r desc_a_crash2b, message = FALSE, results='asis', warning=FALSE, , echo=FALSE}
a_crash2 %>%
  dplyr::select(all_of(pyhiso_vars)) %>%
  Hmisc::describe(descript = "Physiological measurements") %>%
  Hmisc::html(size = 80)
```

### Characteristics of injury

```{r desc_a_crash2c, message = FALSE, results='asis', warning=FALSE, echo=FALSE}
a_crash2 %>%
  dplyr::select(all_of(injury_vars)) %>%
  Hmisc::describe(descript = "Characteristics of injury") %>%
  Hmisc::html(size = 80)
```

## Categorical variables

We now provide a closer visual examination of the categorical predictors. 

```{r catplot, message=FALSE, warning =FALSE , echo=FALSE}
a_crash2 %>%
  dplyr::select(sex, injurytype) %>%
  dplyr::mutate_all(forcats::as_factor) %>%   
  dplyr::mutate_all(forcats::fct_explicit_na, "NA") %>%
  tidyr::pivot_longer(
    dplyr::everything(),
    names_to = "var",
    values_to = "value",
    values_drop_na = FALSE
  ) %>%
  dplyr::group_by(var, value) %>%
  dplyr::summarize(N = n()) %>%
  dplyr::mutate(
    freq = N / sum(N),
    pct = round((freq * 100), 1),
    axis_lab = paste0(value, ' ', '(N = ', N, ')'),
    var_label = case_when(var == "sex" ~ "Sex",
                          var == "injurytype" ~ "Injury type")
  ) %>%
  ggplot(aes(
    x = reorder(axis_lab, pct),
    y = pct,
    label = pct
  )) +
  geom_text(nudge_y = 7) +
  geom_pointrange(aes(ymin = 0, ymax = pct), alpha = 1, size = 1, color = "grey") +
  geom_point(color = "firebrick2",
             alpha = 0.6,
             size = 3) +
  ylab("Percentage (%)") +
  scale_y_continuous(limits = c(0, 100)) +
  labs(caption = "Number of subjects with a non-missing value reported in brackets.\nNA = missing") +
  facet_wrap(~ var_label, ncol = 1, scales = "free_y") +
  coord_flip() +
  theme_minimal(base_size = 12) +
  theme(axis.title.y = element_blank(),
        panel.grid.minor = element_blank())

```

### Categorical ordinal plots

The Glasgow coma score, an ordinal categorical variable, is also displayed separately. 

```{r, message=FALSE, warning =FALSE , echo=FALSE}
a_crash2 %>%
  dplyr::select(gcs) %>%
  dplyr::mutate(gcs = as_factor(gcs)) %>%
  tidyr::pivot_longer(
    dplyr::everything(),
    names_to = "var",
    values_to = "value",
    values_drop_na = FALSE
  ) %>%
  dplyr::group_by(var, value) %>%
  dplyr::summarize(N = n()) %>%
  dplyr::mutate(
    freq = N / sum(N),
    pct = round((freq * 100), 1),
    value = as_factor(value),
    axis_lab = as_factor(paste0(value, ' ', '(N = ', N, ')'))
  ) %>%
  ggplot(aes(
    x = axis_lab,
    y = pct,
    label = pct
  )) +
  geom_text(nudge_y = 7) +
  geom_pointrange(aes(ymin = 0, ymax = pct), alpha = 1, size = 1, color = "grey") +
  geom_point(color = "firebrick2",
             alpha = 0.6,
             size = 3) +
  ylab("Percentage (%)") +
  scale_y_continuous(limits = c(0, 100)) +
  labs(title = "Glasgow coma score (point scale)",
    caption = "Number of subjects with a non-missing value reported in brackets.\nNA = missing") +
  coord_flip() +
  theme_minimal(base_size = 12) +
  theme(axis.title.y = element_blank(),
        panel.grid.minor = element_blank())
```



## Continuous variables 

A closer visual examination of continuous predictors. 

```{r, message =FALSE, warning =FALSE , echo=FALSE, fig.width=9}

a_crash2 %>%
  dplyr::select(age, sbp, rr, cc, hr, injurytime) %>%
  dplyr::mutate() %>%
  dplyr::mutate_all(as.numeric) %>%
  tidyr::pivot_longer(dplyr::everything(),
                      names_to = "var",
                      values_to = "value") %>%
  dplyr::mutate(
    var_label = case_when(
      var == "age" ~ "Age (years)",
      var == "sbp" ~ "Systolic blood pressure (mmHg)",
      var == "hr" ~  "Heart rate (1/min)",
      var ==  "rr" ~ "Respiratory rate (1/min)",
      var ==  "cc" ~ "Central capillary refill time (seconds)", 
      var == "injurytime" ~ "Injury time (days)"
    )) %>%
  ggplot(aes(value)) +
  ylab("Number of subjects") +
  geom_histogram(
    binwidth = 0.5,
    center = 0,
    alpha = 0.6,
    fill = "firebrick2"
  ) +
  facet_wrap(~ var_label, scales = "free", ncol = 2) +
  theme_minimal() +
  theme(axis.title.x = element_blank(),
        panel.grid.minor = element_blank())

```


There is evidence of digit preference. Explore further with targeted summaries. A more detailed univariate summaries for the variables of interest are also provided below. 

### Age

```{r, fig.width=8, fig.height=3.5, fig.cap= "Distribution of subject age [years]" , echo=FALSE}
## This is a function the plots a strip plot, histogram and boxplot, including five number summary. 
ida_plot_univar(a_crash2, "age")
```

Five patients under the age of 17, the inclusion criteria for the study, with one patient aged 1. 

### Blood pressure

```{r, fig.width=8, fig.height=3.5, fig.cap= "Distribution of SBP" , echo=FALSE}
ida_plot_univar(a_crash2, "sbp", bin_width = 0.7)
```


### Respiratory rate


```{r, fig.width=8, fig.height=3.5, fig.cap= "Distribution of respiratory rate" , echo=FALSE}
ida_plot_univar(a_crash2, "rr", n_dodge = 2)
```


### Heart rate


```{r, fig.width=8, fig.height=3.5, fig.cap= "Distribution of heart rate", echo=FALSE}
ida_plot_univar(a_crash2, "hr")
```


### Central capillary refill time


```{r, fig.width=8, fig.height=3.5, fig.cap= "Distribution of Central capillary refill time", echo=FALSE}
ida_plot_univar(a_crash2, "cc", bin_width = 0.5)
```


### Hours since injury


```{r, fig.width=8, fig.height=3.5, fig.cap= "Distribution of hours since injury", echo=FALSE}
ida_plot_univar(a_crash2, "injurytime", n_dodge = 3, bin_width = 0.25)
```



## Section session info

```{r, warning=FALSE, message=FALSE, echo=FALSE}
sessionInfo()
```


