
#  Multivariable/variate distributions

```{r, echo = FALSE, message = FALSE, warning = FALSE }
library(here)
library(tidyverse)
library(Hmisc)
library(arsenal)

## Read data 
load(here::here("data", "a_crash2.rda"))
```



## Summary by sex

```{r, results="asis", message=FALSE, warning =FALSE , echo=FALSE}
crs.char <-
  tableby(
    sex ~ age + hr + rr + sbp ,
    data = a_crash2,
    test = FALSE,
    total = FALSE,
    numeric.stats = c("median","mean", "sd", "q1q3", "range", "Nmiss2"), 
    digits = 1, digits.p = 2, digits.pct = 1
  )

summary(crs.char, title = 'Patient characteristics') 
```

```{r, results="asis", message=FALSE, warning =FALSE , echo=FALSE}
s <-
  summaryM(
    age + hr + rr + sbp + cc + gcs + injurytime + injurytype ~ sex,
    data = a_crash2,
    overall = FALSE,
    test = FALSE
  )


html(
  s,
  caption = 'Baseline characteristics by sex',
  exclude1 = TRUE,
  npct = 'both',
  digits = 3,
  prmsd = TRUE,
  brmsd = TRUE,
  msdsize = mu$smaller2
)
```







## Continuous variables by sex

### Distribution of age by sex

```{r, echo = FALSE, message = FALSE, fig.cap= "Distribution of age by sex"}
# only plot observatons with a non-missing value for Sex
a_crash2 %>%
  filter(!(is.na(sex))) %>%
  with(., histboxp(
    x = age,
    group = sex,
    sd = TRUE,
    bins = 200
  ))


```

### Distribution of systolic blood pressure by sex


```{r, echo = FALSE, message = FALSE, fig.cap= "Distribution of systolic blood pressure by sex"}
a_crash2 %>%
  filter(!(is.na(sex))) %>%
  with(., histboxp(
    x = sbp,
    group = sex,
    sd = TRUE,
    bins = 200
  ))
```


### Distribution of heart rate by sex

```{r, echo = FALSE, message = FALSE, fig.cap= "Distribution of heart rate by sex" }
a_crash2 %>%
  filter(!(is.na(sex))) %>%
  with(., histboxp(
    x = hr,
    group = sex,
    sd = TRUE,
    bins = 200
  ))
```


### Distribution of respiratory rate by sex

```{r, echo = FALSE, message = FALSE, fig.cap= "Distribution of respiratory rate by sex" }
a_crash2 %>%
  filter(!(is.na(sex))) %>%
  with(., histboxp(
    x =  rr,
    group = sex,
    sd = TRUE,
    bins = 200
  ))
```


### Distribution of central capillary refille time by sex

```{r, echo = FALSE, message = FALSE, fig.cap= "Distribution of centrail capillary refille time by sex" }
a_crash2 %>%
  filter(!(is.na(sex))) %>%
  with(., histboxp(
    x =  cc,
    group = sex,
    sd = TRUE,
    bins = 200
  ))
```


## Age 

### Continuous 

```{r}

a_crash2 %>% dplyr::filter(!is.na(sbp)) %>% tally()
a_crash2 %>% dplyr::filter(is.na(sbp)) %>% tally()

bigN <- a_crash2 %>% dplyr::filter(!is.na(sbp) & !is.na(age)) %>% tally()
n_miss <- a_crash2 %>% dplyr::filter(is.na(sbp) | is.na(age)) %>% tally()

title <-
  paste0("Plot of ", Hmisc::label(a_crash2$age), " and ", Hmisc::label(a_crash2$sbp))

caption <-
  paste0(
    "n = ",
    bigN,
    " subjects displayed.\n",
    n_miss,
    " subjects with a missing value in at least one of the variables."
  )


x_axis <- paste0(Hmisc::label(a_crash2$age), " [", Hmisc::units(a_crash2$age), "]")
y_axis <- paste0(Hmisc::label(a_crash2$sbp), " [", Hmisc::units(a_crash2$sbp), "]")


p1 <- a_crash2 %>%
  dplyr::filter(!is.na(sbp) & !is.na(age)) %>%
  mutate(sbp = as.numeric(sbp), 
         age = as.numeric(age)) %>%
  ggplot(aes(x = sbp, y = age)) +
  ylab(x_axis) +
  xlab(y_axis) +
  labs(
    title = title,
    caption = caption
  ) +
  geom_point(shape = 16, #size = 0.5,
             alpha = 0.5,
             color = "firebrick2") +
  geom_rug() +
  theme_minimal()

p1

```

### Continuous

```{r}
p1 <- a_crash2 %>%
  filter(!is.na(sbp) & !is.na(age)) %>%
  mutate(sbp = as.numeric(sbp), 
         age = as.numeric(age)) %>%
  ggplot(aes(x = sbp, y = age)) +
  geom_point(shape = 16, size = 0.5,
             alpha = 0.5,
             color = "firebrick2") +
  geom_rug() +
  theme_minimal()

p2 <- a_crash2 %>%
  filter(!is.na(sbp) & !is.na(hr)) %>%
  mutate(sbp = as.numeric(sbp), 
         age = as.numeric(hr)) %>%
  ggplot(aes(x = sbp, y = hr)) +
  geom_point(shape = 16, size = 0.5,
             alpha = 0.5,
             color = "firebrick2") +
  geom_rug() +
  theme_minimal()


p3 <- a_crash2 %>%
  filter(!is.na(sbp) & !is.na(rr)) %>%
  mutate(sbp = as.numeric(sbp), 
         age = as.numeric(rr)) %>%
  ggplot(aes(x = sbp, y = rr)) +
  geom_point(shape = 16, size = 0.5,
             alpha = 0.5,
             color = "firebrick2") +
  geom_rug() +
  theme_minimal()


p4 <- a_crash2 %>%
  filter(!is.na(hr) & !is.na(age)) %>%
  mutate(sbp = as.numeric(hr), 
         age = as.numeric(age)) %>%
  ggplot(aes(x = hr, y = age)) +
  geom_point(shape = 16, size = 0.5,
             alpha = 0.5,
             color = "firebrick2") +
  geom_rug() +
  theme_minimal()

```


### Continuous3 

```{r}

library(patchwork)
(p1 + p2 ) / (p3 + p4)
```


## Scatter plots with a third or fourth variable

Scatter plot of age and RR by sex and injury type.


```{r}
ggplot(a_crash2, aes(
  y = age,
  x = rr
)) +
    geom_point(shape = 16, size = 0.5,
             alpha = 0.5,
             color = "firebrick2") +
  geom_rug() +
  facet_grid(sex ~ injurytype) +
  theme_minimal()
```


Scatter plot of SBP and RR by sex and injury type.

```{r}
ggplot(a_crash2, aes(
  y = sbp,
  x = rr
)) +
    geom_point(shape = 16, size = 0.5,
             alpha = 0.5,
             color = "firebrick2") +
  geom_rug() +
  facet_grid(sex ~ injurytype) +
  theme_minimal()
```




## Session info

```{r}
sessionInfo()
```
