# CRASH-2 {-}


```{r, echo=FALSE, warning=FALSE, message=FALSE, echo=FALSE}
## Load libraries for this chapter
library(here)
library(tidyverse)
library(Hmisc)
```

## Introduction to CRASH-2

**Description:** Clinical Randomisation of an Antifibrinolyticin Significant Haemorrhage(CRASH-2)  was a large randomised placebo controlled trial among trauma patients with, or at risk of, significant haemorrhage, of the effects of antifibrinolytic treatment on death and transfusion requirement. The study is described at [the original trial website](http://crash2.lshtm.ac.uk/). A public version of the data set is found at a [repository of public data sets](http://biostat.mc.vanderbilt.edu/wiki/Main/DataSets) hosted by the Vanderbilt University's Department of Biostatistics (Prof. Frank Harrell Jr.).

A prediction model was devloped and valdiated based on this data set in "Predicting early death in patients with traumatic bleeding"  [Perel et al, BMJ 2012](https://doi.org/10.1136/bmj.e5166), [supplement available at](https://www.bmj.com/highwire/filestream/634478/field_highwire_adjunct_files/0/perp003158.ww1_default.pdf).

The data set includes 20,207 patients and 44 variables. 



## Crash2 dataset contents

### Source dataset 

Display the source dataset contents. The dataset is in the **data-raw** folder of the project directory. 


**TODO: Move the contents of the original data set to an appendix? IS it relevant for us?   No, it looks good here-MH**

```{r contents_crash2,  warning=FALSE, message=FALSE, echo=FALSE, results='asis'} 
load(here::here("data-raw", "crash2.rda"))

Hmisc::html(Hmisc::contents(crash2),
            maxlevels = 10,
            levelType = 'table')

```


### Updated analysis dataset

Additional meta-data is added to the original *source* data set. We write this new modified data set back to the **data** folder after adding additional meta-data for the following variables:

* **age** - add label "Age" and unit "years". 
* **injury time** - add unit "hours". 
* **total Glasgow coma score** - add unit "points". 

**TODO:** 

* Do we want to select patients at this point or leave this for the analysis phase?
* Do we also want to do a selection of variables here to take in to the IDA phase? i.e. drop variables we do not check in IDA?

As a cross check we display the contents again to ensure the additional data is added, and then write back the changes to the data folder in the file "data/a_crash2.rds". 

```{r contents_acrash2, warning=FALSE, message=FALSE, results='asis'} 

## Complete metadata by adding missing labels. 
## Generate adervived dataset stored in data as we are adding to the oirginal source dataset obtained. 

## select candidate predictor variables. -- See SAP
crash2_subset <-
  crash2 %>%
  dplyr::select(
    entryid,        # patient identifer
    trandomised,    # date of randomisation 
    ddeath,         # date of death
    age,            # Age (`age`, years)
    sex,            # Sex (`sex`, male or female)
    sbp,            # Systolic blood pressure (`sbp`, mmHg)
    hr,             # Heart rate (`hr`, 1/min)
    rr,             # Respiratory rate (`rr`, 1/min)
    gcs,            # Glasgow coma score (`gcs`, points)
    cc,             # Central capillary refill time (`cc`, seconds)
    injurytime,     # Time since injury (`injurytime`, hours)
    injurytype      # Type of injury (`injurytype`, 'blunt', 'penetrating' or 'blunt and penetrating')
  )


## Complete metadata by adding missing labels.
a_crash2 <- Hmisc::upData(
  crash2_subset,
  labels = c(
    age = 'Age',
    sex = "Sex", 
    injurytype = "Injury type", 
    time2death = "Time from randomization to day of death"),
  units = c(
    age = "years",
    injurytime = "hours",
    gcs = "points", 
    time2death = "days"
  )
)


## Derivie outcome variable
a_crash2$time2death <-
  as.numeric(as.Date(a_crash2$ddeath) - as.Date(a_crash2$trandomised))

a_crash2$earlydeath[!is.na(a_crash2$time2death)] <-
  (a_crash2$time2death[!is.na(a_crash2$time2death)] <= 28) + 0

# +0 to transform it from TRUE/FALSE to 1/0
# NA in time2death means alive at day 28
a_crash2$earlydeath[is.na(a_crash2$time2death)] <- 0    

## Add meta data
a_crash2 <- Hmisc::upData(a_crash2 ,
               labels = c(earlydeath = 'Death within 28 days from injury'))


## Display contents
Hmisc::html(Hmisc::contents(a_crash2),
            maxlevels = 10,
            levelType = 'table')

## Write to data folder
save(a_crash2, file = here::here("data", "a_crash2.rda"))
```
