# Multivariate distributions

```{r setup, echo=FALSE}

knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  fig.align = 'center'
)
```


This section reports a series of multivariate summaries of the NHANES dataset.

```{r}
library(here)
library(tidyverse)
library(plotly)
library(Hmisc)
library(gtsummary)
library(corrplot)
library(ggplot2)
library(naniar)
library(gridExtra)


# Define sets of predictors by type - helpful for selection
#mortstat  (one year mortality)
demog_vars <- c("age", "gender", "educationadult","bmi", "smokecigs" , "drinkstatus")
life_vars <- c("bmi", "smokecigs" , "drinkstatus")
health_vars <- c("diabetes", "chf", "cancer" , "stroke")
physio_vars <- c("sys", "lbxtc",  "lbdhdd")
pa_vars <- c("mobilityproblem", "tac" ,"tlac" ,"mvpa" ,  "wt"  )


## Read data 
#load(here::here("data", "a_nhanes.rda"))
#load(here::here("data", "nhanesdat.rdata"))

load(here::here("a_nhanes.rda"))
load(here::here("~/Documents/TG3/IDA_Regression/github_nhanes/nhanesdat.rdata"))

a_nhanes$agegrp<-cut(a_nhanes$age, breaks=c(0, 45, 60, 75, Inf), labels=c("30-44", "45-59", "60-74", "75+"), right=FALSE)


```


## Overview

### Variable correlation

```{r}

corrs <- a_nhanes %>%
  dplyr::select(age, gender, bmi, sys,lbxtc,  lbdhdd, smokecigs , drinkstatus, mortstat ,diabetes, chf, cancer , stroke ) %>%
  filter(complete.cases(.)) %>%
  dplyr::mutate_all(as.numeric)
M <- cor(corrs)
col <- colorRampPalette(c("#BB4444", "#EE9988", "#FFFFFF", "#77AADD", "#4477AA"))
corrplot(M, method = "color", col = col(200),
         type = "upper", order = "hclust", number.cex = .7,
         addCoef.col = "black", # Add coefficient of correlation
         tl.col = "black", tl.srt = 90, # Text label color and rotation
         # hide correlation coefficient on the principal diagonal
         diag = FALSE)
```


Correlations of the physical activity variables (outcome)
```{r}
corrs <- a_nhanes %>%
  dplyr::select(mobilityproblem, tac ,tlac ,mvpa ,  wt  ) %>%
  filter(complete.cases(.)) %>%
  dplyr::mutate_all(as.numeric)
M <- cor(corrs)
col <- colorRampPalette(c("#BB4444", "#EE9988", "#FFFFFF", "#77AADD", "#4477AA"))
corrplot(M, method = "color", col = col(200),
         type = "upper", order = "hclust", number.cex = .7,
         addCoef.col = "black", # Add coefficient of correlation
         tl.col = "black", tl.srt = 90, # Text label color and rotation
         # hide correlation coefficient on the principal diagonal
         diag = FALSE)




```
 
### Variable clustering

*Variable clustering is used for assessing collinearity, redundancy, and for separating variables into clusters that can be scored as a single variable, thus resulting in data reduction.*

```{r}
Hmisc::varclus( ~ age +gender+ bmi+ sys+ lbxtc+  lbdhdd+ smokecigs + drinkstatus+ mortstat + diabetes+ chf+ cancer + stroke, data = a_nhanes)
```

Plot associations. 

```{r}
plot(Hmisc::varclus( ~ age +gender+ bmi+ sys+ lbxtc+  lbdhdd+ smokecigs +  mortstat + diabetes+ chf+ cancer + stroke, data = a_nhanes))
```


### Variable redundancy 

Redundancy analysis of predictor variables. 

```{r}
Hmisc::redun( ~ age +gender+ bmi+ sys+ lbxtc+  lbdhdd+ smokecigs + drinkstatus+ mortstat + diabetes+ chf+ cancer + stroke  , data = a_nhanes)
```


##  Summary reports by age and gender

### Distribution of age by gender

```{r,  fig.cap= "Distribution of age by gender"}
# only plot observatons with a non-missing value for Sex
a_nhanes %>%
  filter(!(is.na(gender))) %>%
  with(., histboxp(
    x = age,
    group = gender,
    sd = TRUE,
    bins = 200
  ))
```


## Summary report by age group and gender

### Summary report by gender

```{r, results="asis"}
s <-
  Hmisc::summaryM(
     age+bmi+ educationadult+ sys+ lbxtc+  lbdhdd+ smokecigs + drinkstatus+ mortstat + diabetes+ chf+ cancer + stroke ~ gender,
    data = a_nhanes,
    overall = FALSE,
    test = FALSE
  )
Hmisc::html(
  s,
  caption = 'Baseline characteristics by gender',
  exclude1 = TRUE,
  npct = 'both',
  digits = 3,
  prmsd = TRUE,
  brmsd = TRUE,
  msdsize = mu$smaller2
)

```


### Summary report by age group for men

```{r}

sub<-a_nhanes[a_nhanes$gender=="Male",]


s <-
  Hmisc::summaryM(
     bmi+ educationadult+ sys+ lbxtc+  lbdhdd+ smokecigs + drinkstatus + diabetes+ chf+ cancer + stroke ~ agegrp,
    data = sub,
    overall = FALSE,
    test = FALSE
  )
Hmisc::html(
  s,
  caption = 'Baseline characteristics by gae group for men',
  exclude1 = TRUE,
  npct = 'both',
  digits = 3,
  prmsd = TRUE,
  brmsd = TRUE,
  msdsize = mu$smaller2
)

```



```{r, eval=FALSE, include=FALSE}
sub<-a_nhanes[a_nhanes$gender=="Male",]

sub[,c("agegrp", "educationadult", life_vars, physio_vars, health_vars)] %>% 
    tbl_summary(
    by = agegrp,
    statistic = list(all_continuous() ~ "{median} ({p25}, {p75})",
                     all_categorical() ~ "{n} / {N} ({p}%)"),
    digits = all_continuous() ~ 2,
    label = list( smokecigs ~ "Smoking status",
                  drinkstatus ~ "Alcohol consumption",
                  educationadult ~ "Education level",
                  lbxtc ~ "Total cholesterol",
                  lbdhdd ~ "HDL"),
    missing_text = "Missing"
  ) 

```



### Summary report by age group  for women


```{r}

sub<-a_nhanes[a_nhanes$gender=="Female",]

s <-
  Hmisc::summaryM(
     bmi+ educationadult+ sys+ lbxtc+  lbdhdd+ smokecigs + drinkstatus + diabetes+ chf+ cancer + stroke ~ agegrp,
    data = sub,
    overall = FALSE,
    test = FALSE
  )
Hmisc::html(
  s,
  caption = 'Baseline characteristics by gae group for men',
  exclude1 = TRUE,
  npct = 'both',
  digits = 3,
  prmsd = TRUE,
  brmsd = TRUE,
  msdsize = mu$smaller2
)

```




## Continuous variables by age and gender

### Distribution of systolic blood pressure

```{r}

g<-ggplot(a_nhanes, 
       aes(x = age, 
           y = sys)) + 
  geom_point(alpha=1/10)+
  #geom_miss_point() + 
  facet_wrap(~gender)+ 
  theme_bw() + theme(legend.title = element_blank()) + ylab("systolic blood pressure")

g

```


### Distribution of cholesterol

```{r}

g<-ggplot(a_nhanes, 
       aes(x = age, 
           y = lbxtc)) + 
  geom_point(aes(color = gender))+
  #geom_miss_point() + 
  #facet_wrap(~gender)+ 
  theme_bw() + theme(legend.title = element_blank()) 

g + ylab("total cholesterol")

```


```{r, results='asis', message =FALSE, warning =FALSE , echo=FALSE}

g<-ggplot(a_nhanes, 
       aes(x = age, 
           y = lbdhdd)) + 
  geom_point(aes(color = gender))+
  #geom_miss_point() + 
  #facet_wrap(~gender)+ 
  theme_bw() + theme(legend.title = element_blank()) 
g + ylab("HDL")

```

### Distribution of BMI

```{r}

g<-ggplot(a_nhanes, 
       aes(x = age, 
           y = bmi)) + 
  geom_point(aes(color = gender))+
  #geom_miss_point() + 
  #facet_wrap(~gender)+ 
  theme_bw() + theme(legend.title = element_blank()) + ylab("systolic blood pressure")
g

```

### Distribution of wear time

```{r}

g<-ggplot(a_nhanes, 
       aes(x = age, 
           y = wt)) + 
  geom_point(alpha=1/10)+
  #geom_miss_point() + 
  facet_wrap(~gender)+ 
  theme_bw() + theme(legend.title = element_blank()) + ylab("wear time")

g

```


## Physical activity data

### Distribution of MVPA 

```{r}
g1<-ggplot(a_nhanes, aes(wt, mvpa)) +
  geom_point(alpha=1/10) +
  geom_smooth()

g2<-ggplot(a_nhanes, aes(wt, log(1+mvpa))) +
  geom_point(alpha=1/10) +
  geom_smooth()

g3<-ggplot(a_nhanes, aes(tlac, mvpa)) +
  geom_point(alpha=1/10) +
  geom_smooth()

g4<-ggplot(a_nhanes, aes(tlac, log(1+mvpa))) +
  geom_point(alpha=1/10) +
  geom_smooth()

grid.arrange(g1, g2,g3, g4,  nrow = 2) 
```


### Distribution of MVPA and Total log activity count by time of day

```{r}


#12AM-2AM, 2AM-4AM, 4AM-6AM

g1<-ggplot(nhanesdat, aes(tlac.1, mvpa)) +
  geom_point(alpha=1/10) +
  geom_smooth() +xlim(0,800) +xlab("tlac 12AM-2AM")

g2<-ggplot(nhanesdat, aes(tlac.2, mvpa)) +
  geom_point(alpha=1/10) +
  geom_smooth() +xlim(0,800) +xlab("tlac 2AM-4AM")

g3<-ggplot(nhanesdat, aes(tlac.3, mvpa)) +
  geom_point(alpha=1/10) +
  geom_smooth()+xlim(0,800) +xlab("tlac 4AM-6AM")

g4<-ggplot(nhanesdat, aes(tlac.4, mvpa)) +
  geom_point(alpha=1/10) +
  geom_smooth() +xlim(0,800)+xlab("tlac 6AM-8AM")

g5<-ggplot(nhanesdat, aes(tlac.5, mvpa)) +
  geom_point(alpha=1/10) +
  geom_smooth() +xlim(0,800)+xlab("tlac 8AM-10AM")

g6<-ggplot(nhanesdat, aes(tlac.6, mvpa)) +
  geom_point(alpha=1/10) +
  geom_smooth() +xlim(0,800)+xlab("tlac 10AM-12PM")

g7<-ggplot(nhanesdat, aes(tlac.7, mvpa)) +
  geom_point(alpha=1/10) +
  geom_smooth() +xlim(0,800)+xlab("tlac 12PM-2PM")

g8<-ggplot(nhanesdat, aes(tlac.8, mvpa)) +
  geom_point(alpha=1/10) +
  geom_smooth() +xlim(0,800)+xlab("tlac 2PM-4PM")


g9<-ggplot(nhanesdat, aes(tlac.9, mvpa)) +
  geom_point(alpha=1/10) +
  geom_smooth() +xlim(0,800)+xlab("tlac 4PM-6PM")

g10<-ggplot(nhanesdat, aes(tlac.10, mvpa)) +
  geom_point(alpha=1/10) +
  geom_smooth() +xlim(0,800)+xlab("tlac 6PM-8PM")

g11<-ggplot(nhanesdat, aes(tlac.11, mvpa)) +
  geom_point(alpha=1/10) +
  geom_smooth() +xlim(0,800)+xlab("tlac 8PM-10PM")

g12<-ggplot(nhanesdat, aes(tlac.12, mvpa)) +
  geom_point(alpha=1/10) +
  geom_smooth() +xlim(0,800)+xlab("tlac 10PM-12AM")


grid.arrange(g1, g2,g3, g4,g5,g6,g7,g8,g9,g10,g11,g12,  nrow = 4) 


```


## Section session info

```{r, warning=FALSE, message=FALSE, echo=FALSE}
sessionInfo()
```
