# Missing data


```{r, echo=FALSE, warning=FALSE, message=FALSE, echo=FALSE}
## Load libraries
library(here)
library(tidyverse)
library(Hmisc)
library(naniar) ## for missing data 
library(gt)
library(kableExtra)
library(DT)
library(arsenal)
#load(here::here("data", "a_nhanes.rda"))

#load(here::here("a_nhanes.rda"))
load(here::here("data", "a_nhanes.rda"))
```


## Per variable missingness

Number and percentage of missing. 

```{r, message =FALSE, warning =FALSE , echo=FALSE}
# devtools::install_github("rstudio/gt")
a_nhanes %>%
  select(-seqn) %>%
  miss_var_summary() %>%
  gt::gt() %>%
  gt::cols_label(
    variable = md("**Variable**"),
    n_miss = md("**Missing (count)**"),
    pct_miss = md("**Missing (%)**")
  ) %>%
  gt::fmt_number(
    columns = vars(pct_miss),
    decimals = 2
  )
```


```{r, message =FALSE, warning =FALSE , echo=FALSE}
a_nhanes %>%
  select(-seqn) %>%
  gg_miss_var(show_pct = TRUE)
```

## Variable summaries for complete vs incomplete cases

```{r, results='asis', message =FALSE, warning =FALSE , echo=FALSE}
a_nhanes$mvpamiss<-"incomplete"
idx<-which(is.na(a_nhanes$mvpa))
a_nhanes$mvpamiss[idx]<-"complete"

demog_vars <- c("age", "gender", "educationadult","bmi", "smokecigs" , "drinkstatus")
life_vars <- c("bmi", "smokecigs" , "drinkstatus")
health_vars <- c("mortstat" ,"diabetes", "chf", "cancer" , "stroke")
physio_vars <- c("sys","lbxtc",  "lbdhdd")
pa_vars <- c("mobilityproblem", "tac" ,"tlac" ,"mvpa" ,  "wt" , "satp" , "astp" )

vars<-c(demog_vars, health_vars,physio_vars, "mobilityproblem", "tlac", "wt", "mvpamiss" )

nhanes_complete<-tableby(mvpamiss ~ ., data=a_nhanes[,vars], test=TRUE, numeric.stats=c("median","q1q3", "range"))

# need results='asis' above for this to look good in the output
summary(nhanes_complete, title = "Participant characteristics by missing status of MVPA")

```

It appears that participants with incomplete physical activity data are older.


## Missingness patterns over variables

```{r, message =FALSE, warning =FALSE , echo=FALSE}
library(VIM)
aggr(a_nhanes[,vars],numbers=TRUE)
```


```{r, message =FALSE, warning =FALSE , echo=FALSE}
a_nhanes %>%
  select(-seqn) %>%
  naniar::vis_miss(sort_miss = TRUE, show_perc_col = TRUE) 
```

```{r, message =FALSE, warning =FALSE , echo=FALSE}
a_nhanes %>%
  select(-seqn) %>%
  gg_miss_case()
```


```{r, message =FALSE, warning =FALSE , echo=FALSE}
a_nhanes %>%
  select(mvpa, tlac, drinkstatus, smokecigs, sys, lbxtc,lbdhdd, bmi, educationadult, ) %>%
  gg_miss_upset()
```




## (In)complete cases

This section presents patients with a least one missing value. First we list out patients with at least one missing value in a filterable table. 

```{r, message =FALSE, warning =FALSE , echo=FALSE}
cc <-
  a_nhanes %>%  
  select(-seqn,-mvpa) %>% 
  filter(!complete.cases(.))
DT::datatable(cc)
```

Then we report the pattern of missing for this set of patients. 

```{r, message =FALSE, warning =FALSE , echo=FALSE}
cc %>% gg_miss_upset()
```



## Section session info

```{r, message =FALSE, warning =FALSE , echo=FALSE}
sessionInfo()
```