# NHANES {-}


```{r, echo=FALSE, warning=FALSE, message=FALSE, echo=FALSE}
## Load libraries for this chapter
library(here)
library(tidyverse)
library(Hmisc)
```

# Introduction to NHANES

Since a key principle of IDA is not to touch the research questions, before IDA commences the research aim and statistical analysis plan need to be in place. 
IDA may lead to an update or refinement of the analysis plan. To demonstrate the workflow and content of IDA, we created a hypothetical research aim and corresponding statistical analysis plan, which is described in more detail in the section [nhanes_SAP.Rmd](nhanes_SAP.Rmd).

**Hypothetical research aim for IDA** is to 
develop a multivariable model for MVPA (minutes of moderate/vigorous physical activity) with primary aim of variable selection to predict MVPA and secondary aim to study the role of systolic blood pressure in addition to variables identified.

The preparation of the data was based on "Organizing and Analyzing the Activity Data in NHANES"
  [Leroux et al, Statistics in Biosciences 2019](https://doi.org/10.1007/s12561-018-09229-9). 
  
The inclusion criteria are in line with the choices for the prediction model of 5 year mortality presented by [Smirnova et al, J Gerontol A Biol Sci Med Sci 2020](https://doi.org/10.1093/gerona/glz193).


## NHANES Description

The National Health and Nutrition Examination Survey (NHANES) is a program of studies designed to assess the health and nutritional status of adults and children in the United States. The survey examines a nationally representative sample of about 5,000 persons each year. [Link to CDC  NHANES website](https://www.cdc.gov/nchs/nhanes/about_nhanes.htm)

The NHANES study contains objectively measured physical activity data collected using hip-worn accelerometers from multiple cohorts. High quality processed activity data combined with mortality and demographic information can be downloaded and used in R with code from Andrew Leroux (https://andrew-leroux.github.io/rnhanesdata/articles).


The NHANES 2003–2004 and 2005–2006 have a total of 14,631 participants with accelerometry data. Similarly to Smirnova et al 2020 we focus on participants for ages 50 to 85 at the time they wore the accelerometer. 




## NHANES dataset contents

### Source dataset 

We refer to the source data set as the dataset available online [here](.... Github .....)

Display the source dataset contents. This dataset is in the **data-raw** folder of the project directory. 

```{r contents_nhanes,  warning=FALSE, message=FALSE, echo=FALSE, results='asis'} 
## Load the dataset from the data-raw folder. 
## Alternatively we could load directly from the url 

#load(here::here("data-raw", "nhanesdat.rdata"))

load(here::here( "nhanesdat.rdata"))
Hmisc::html(Hmisc::contents(nhanesdat),
            maxlevels = 10,
            levelType = 'table')
```


### Updated analysis dataset

Additional meta-data is added to the original *source* data set. We write this new modified data set back to the **data** folder after adding additional meta-data for the following variables:

* **sys** - add label "Systolic Blood pressure" and unit "mg/dl". 


Set missing 'drinkstatus' to NA
```{r}
indx<-which(nhanesdat$drinkstatus=='Missing alcohol')
nhanesdat$drinkstatus[indx]<-NA

nhanesdat <- nhanesdat %>% 
        mutate(mortstat = factor(mortstat, levels = 0:1, labels = c("alive", "dead")))
```

At the stage we select the variables of interest to take in to the IDA phase by dropping variables we do not check in IDA.

As a cross check we display the contents again to ensure the additional data is added, and then write back the changes to the data folder in the file "data/a_nhanes.rda". 

```{r labels_nhanes, warning=FALSE, message=FALSE, echo=FALSE, results='asis'} 
## Complete metadata by adding missing labels. 
## Generate adervived dataset stored in data as we are adding to the oirginal source dataset obtained. 
## select candidate predictor variables. -- See SAP
nhanes_subset <-
  nhanesdat %>%
  dplyr::select(
    seqn,           # respondent sequence number
    age,            # age at examination, years
    gender,          # gender, Male, Female
    permth.exm,      # Person Months of Follow-up from MEC/Exam Date, follow-up time [years] =permth.exm/12
    mortstat,        # Final mortality status, 0, 1, NAs for individuals with follow up less than 5 years and alive)
    educationadult,  # education level ("Less than high school", "High school", "More than high school")
    smokecigs,       # smoking status (Current, Former, Never) 
    drinkstatus,     # alcohol consumption (Non-Drinker, Moderate Drinker, Heavy Drinker, Missing alcohol)
    bmi ,            # body mass index , kg/m2
    diabetes,        # diabetes (No-Yes)
    chf ,            # congestive heart failure (No-Yes)
    cancer,           # cancer (No-Yes)
    stroke,           # stroke (No-Yes)
    sys ,             #average systolic blood pressure using the 4 measurements, mmHg 
    lbxtc,           #  Total cholesterol,   mg/dL
    lbdhdd,           # HDL cholesterol, mg/dL 
    mobilityproblem,  # difficulties with mobility ( "No Difficulty", "Any Difficulty")
    tac,               #total activity counts per day
    tlac,              # total log activity count (log(1+activity))
    mvpa,             # total minutes of moderate/vigorous physical activity
    wt,               # total accelerometer wear time
    satp,             #sedentary/sleep/non-wear to active transition probability
    astp              # active to sedentary/sleep/non-wear transition probability
  )

## Complete metadata by adding labels.
a_nhanes <- Hmisc::upData(
  nhanes_subset,
  labels = c(
    sys = 'Systolic blood pressure',
    mvpa = "Moderate or vigorous physical activity "),
  units = c(
    age = "years",
    sys = "mg/dl",
    mvpa = "minutes"
  )
)


## Display contents
Hmisc::html(Hmisc::contents(a_nhanes),
            maxlevels = 10,
            levelType = 'table')
## Write to data folder
#save(a_nhanes, file = here::here("data", "a_nhanes.rda"))

save(a_nhanes, file = here::here("a_nhanes.rda"))
```

## Section session info

```{r, warning=FALSE, message=FALSE, echo=FALSE}
sessionInfo()
```