# Univariate distribution checks

```{r setup, echo=FALSE}

knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  fig.align = 'center'
)
```


This section reports a series of univariate summary checks of the NHANES dataset.

```{r, message =FALSE, warning =FALSE , echo=FALSE}
library(here)
library(tidyverse)
library(Hmisc)


source(here("R", "ida_plot_univar.R"))  ## function to plot univariate summaries. 

## Load the dataset.
load(here::here("data", "a_nhanes.rda"))

# load(here::here("a_nhanes.rda"))


# Define sets of predictors 
vip_vars <-c("age", "gender")
demog_vars <- c("age", "gender", "educationadult")
life_vars <- c("bmi", "smokecigs" , "drinkstatus")
health_vars <- c("diabetes", "chf", "cancer" , "stroke")
physio_vars <- c("sys","lbxtc",  "lbdhdd")
pa_vars <- c( "tac" ,"tlac" ,"mvpa", "wt" , "tlac.1", "tlac.2","tlac.3","tlac.4","tlac.5","tlac.6","tlac.7","tlac.8","tlac.9","tlac.10","tlac.11","tlac.12" )

vars<-c(demog_vars,life_vars,health_vars, physio_vars, "tac" ,"tlac" ,"mvpa", "wt" )


```




## Data set overview

Using the [Hmisc](https://cran.r-project.org/web/packages/Hmisc/) describe function, we provide an overview of the data set. The descriptive report also provides histograms of continuous variables. For ease of scanning the information, we group the report by measurement type. 

### Demographic and lifestyle variables

```{r desc_a_nhanesa,  results='asis'}
a_nhanes %>%
  dplyr::select(all_of(c(demog_vars, life_vars))) %>%
  Hmisc::describe(descript = "Demographic and lifestyle variables") %>%
  Hmisc::html(size = 80)
```



### Physiological measurements

```{r desc_a_nhanesc,  results='asis', }
a_nhanes %>%
  dplyr::select(all_of(physio_vars)) %>%
  Hmisc::describe(descript = "Physiological measurements") %>%
  Hmisc::html(size = 80)
```


### Comorbidities

```{r desc_a_nhanesd,  results='asis'}
a_nhanes %>%
  dplyr::select(all_of(health_vars)) %>%
  Hmisc::describe(descript = "Comorbidities") %>%
  Hmisc::html(size = 80)
```

### Physical activity variables

```{r desc_a_nhanesb,  results='asis'}
a_nhanes %>%
  dplyr::select(all_of(pa_vars)) %>%
  Hmisc::describe(descript = "Physical activity") %>%
  Hmisc::html(size = 80)
```


## Categorical variables

We now provide a closer visual examination of the categorical predictors. 

```{r catplot,  fig.height=10}
a_nhanes %>%
  dplyr::select(gender, educationadult,  smokecigs , drinkstatus, health_vars) %>%
  dplyr::mutate_all(forcats::as_factor) %>%   
  dplyr::mutate_all(forcats::fct_explicit_na, "NA") %>%
  tidyr::pivot_longer(
    dplyr::everything(),
    names_to = "var",
    values_to = "value",
    values_drop_na = FALSE
  ) %>%
  dplyr::group_by(var, value) %>%
  dplyr::summarize(N = n()) %>%
  dplyr::mutate(
    freq = N / sum(N),
    pct = round((freq * 100), 1),
    axis_lab = paste0(value, ' ', '(N = ', N, ')'),
    var_label = case_when(var == "gender" ~ "Gender",
                          var == "educationadult" ~ "Education status",
                          var == "smokecigs" ~ "Smoking status",
                          var == "drinkstatus" ~ "Alcohol consumption",
                          var == "diabetes" ~ "Diabetes", 
                          var == "chf" ~ "Congestive heart failure", 
                          var == "cancer"~ "Cancer" , 
                          var == "stroke" ~ "Stroke")
  ) %>%
  ggplot(aes(
    x = reorder(axis_lab, pct),
    y = pct,
    label = pct
  )) +
  geom_text(nudge_y = 7) +
  geom_pointrange(aes(ymin = 0, ymax = pct), alpha = 1, size = 1, color = "grey") +
  geom_point(color = "firebrick2",
             alpha = 0.6,
             size = 3) +
  ylab("Percentage (%)") +
  scale_y_continuous(limits = c(0, 100)) +
  labs(caption = "Number of subjects with a non-missing value reported in brackets.\nNA = missing") +
  facet_wrap(~ var_label, ncol = 1, scales = "free_y") +
  coord_flip() +
  theme_minimal(base_size = 12) +
  theme(axis.title.y = element_blank(),
        panel.grid.minor = element_blank())
```




## Continuous variables 

A closer visual examination of continuous predictors and the outcome variable. 

```{r,  fig.width=9}
a_nhanes %>%
  dplyr::select(mvpa, tlac, age, sys, bmi, wt,  lbxtc, lbdhdd) %>%
  dplyr::mutate() %>%
  dplyr::mutate_all(as.numeric) %>%
  tidyr::pivot_longer(dplyr::everything(),
                      names_to = "var",
                      values_to = "value") %>%
  dplyr::mutate(
    var_label = case_when(
      var == "mvpa" ~ "MVPA (minutes)",
      var == "age" ~ "Age",
      var == "sys" ~ "Systolic blood pressure",
      var == "bmi" ~  "Body mass index",
      var ==  "wt" ~ "wear time (minutes)",
      var ==  "tlac" ~ "total log activity count",
      var == "lbxtc" ~ "cholesterol (total)",   
      var == "lbdhdd" ~ "cholesterol (HDL)"
    )) %>%
  ggplot(aes(value)) +
  ylab("Number of participants") +
  geom_histogram( #aes(y=..density..),
    binwidth = 0.5,
    center = 0,
    alpha = 0.6,
    fill = "firebrick2"
  ) +
  facet_wrap(~ var_label, scales = "free", ncol = 2) +
  theme_minimal() +
  theme(axis.title.x = element_blank(),
        panel.grid.minor = element_blank())
```


There is evidence of influential points in some of the distributions. This is explored further with targeted summaries. A more detailed univariate summaries for the variables of interest are also provided below. 

### Age

```{r, fig.width=8, fig.height=3.5, fig.cap= "Distribution of age" , echo=FALSE}
ida_plot_univar(a_nhanes, "age", bin_width = 0.7)
```


### Blood pressure

```{r, fig.width=8, fig.height=3.5, fig.cap= "Distribution of SBP" }
ida_plot_univar(a_nhanes, "sys", bin_width = 0.7)
```


### Body mass index


```{r, fig.width=8, fig.height=3.5, fig.cap= "Distribution of respiratory rate" }
ida_plot_univar(a_nhanes, "bmi", n_dodge = 2)
```

There is a participant with an unusual high value (130.2). It is possible that this is an entry error (bmi=30.2).





### Outcome= Time of moderate or vigrous physical activity and related variables

```{r, fig.width=8, fig.height=3.5, fig.cap= "Distribution of time of moderate/vigorous activity [minutes]" }
## This is a function the plots a strip plot, histogram and boxplot, including five number summary. 
ida_plot_univar(a_nhanes, "mvpa")
```

After transforming MVPA with $\log_{10}(x+1)$: 

```{r, fig.width=8, fig.height=3.5, fig.cap= "Distribution of log10 (1+MVPA)" }
## This is a function the plots a strip plot, histogram and boxplot, including five number summary. 
a_nhanes <- a_nhanes %>%
            mutate(mvpa_log = log10(1+mvpa))

ida_plot_univar(a_nhanes, "mvpa_log", bin_width = 0.01)
```


```{r, fig.width=8, fig.height=3.5, fig.cap= "Distribution of wear time [minutes]" }
## This is a function the plots a strip plot, histogram and boxplot, including five number summary. 
ida_plot_univar(a_nhanes, "wt")
```


```{r, fig.width=8, fig.height=3.5, fig.cap= "Distribution of log total activity count" }
## This is a function the plots a strip plot, histogram and boxplot, including five number summary. 
ida_plot_univar(a_nhanes, "tlac")
```




## Section session info

```{r, warning=FALSE, message=FALSE, echo=FALSE}
sessionInfo()
```
