# Missing data


```{r, echo=FALSE, warning=FALSE, message=FALSE, echo=FALSE}
## Load libraries
library(here)
library(tidyverse)
library(Hmisc)
library(naniar) ## for missing data 
library(gt)
library(kableExtra)
library(DT)
library(arsenal)
library(VIM)


#load(here::here("data", "a_nhanes.rda"))

load(here::here("a_nhanes.rda"))
a_nhanes <- a_nhanes  %>% drop_na(tlac.1)

a_nhanes<-a_nhanes %>% 
  mutate(alcohol = ifelse(drinkstatus=="Missing alcohol", NA, drinkstatus))

# Define sets of predictors 
vip_vars <-c("age", "gender")
demog_vars <- c("age", "gender", "educationadult")
life_vars <- c("bmi", "smokecigs" , "alcohol")
health_vars <- c("diabetes", "chf", "cancer" , "stroke")
physio_vars <- c("sys","lbxtc",  "lbdhdd")
pa_vars <- c( "tac" ,"tlac" ,"mvpa", "wt" , "tlac.1", "tlac.2","tlac.3","tlac.4","tlac.5","tlac.6","tlac.7","tlac.8","tlac.9","tlac.10","tlac.11","tlac.12" )

vars<-c(demog_vars,life_vars,health_vars, physio_vars, "tac" ,"tlac" ,"mvpa", "wt" )
```


## Per variable missingness

Number and percentage of missing. 

```{r, message =FALSE, warning =FALSE , echo=FALSE}
# devtools::install_github("rstudio/gt")
a_nhanes %>%
  select(all_of(vars)) %>%
  miss_var_summary() %>%
  gt::gt() %>%
  gt::cols_label(
    variable = md("**Variable**"),
    n_miss = md("**Missing (count)**"),
    pct_miss = md("**Missing (%)**")
  ) %>%
  gt::fmt_number(
    columns = vars(pct_miss),
    decimals = 2
  )
```


Investigate for groups of variables:

```{r, message =FALSE, warning =FALSE , echo=FALSE}
a_nhanes %>%
  select() %>%
  mutate(any_variable_missing = ifelse(apply(a_nhanes,1,any_miss),NA,1)) %>%
  mutate(any_demographics_missing = ifelse(apply(a_nhanes[,demog_vars],1,any_miss),NA,1)) %>%
  mutate(any_vip_missing = ifelse(apply(a_nhanes[, vip_vars],1,any_miss),NA,1)) %>%
  mutate(any_lifestyle_missing = ifelse(apply(a_nhanes[,life_vars],1,any_miss),NA,1)) %>%
  mutate(any_health_missing = ifelse(apply(a_nhanes[,health_vars],1,any_miss),NA,1)) %>%
  mutate(any_physact_missing = ifelse(apply(a_nhanes[,pa_vars],1,any_miss),NA,1)) %>%
   mutate(any_lab_missing = ifelse(apply(a_nhanes[,physio_vars],1,any_miss),NA,1)) %>%
  miss_var_summary() %>%
  gt::gt() %>%
    gt::cols_label(
    variable = md("**Variable**"),
    n_miss = md("**Missing (count)**"),
    pct_miss = md("**Missing (%)**")
  ) %>%
  gt::fmt_number(
    columns = vars(pct_miss),
    decimals = 2
  )

```

Overall, 13% are missing when all variables are included in a model. There is no missingness in the VIPs and 7% in life style variables such as smoking and alcohol consumption.


```{r, message =FALSE, warning =FALSE , echo=FALSE}
a_nhanes %>%
  select(vars) %>%
  gg_miss_var(show_pct = TRUE)
```


## Variable summaries for complete vs incomplete cases



```{r, results='asis',message =FALSE, warning =FALSE , echo=FALSE}


#ifelse(apply(a_bact[,c(acute_related_vars)],1,any_miss),NA,1)
a_nhanes$miss<-ifelse(apply(a_nhanes,1,any_miss),"incomplete","complete")



vars<-c(demog_vars, health_vars, life_vars, "tlac", "wt", "miss" )

nhanes_complete<-tableby(miss ~ ., data=a_nhanes[,vars], test=TRUE, numeric.stats=c("median","q1q3", "range"))

# need results='asis' above for this to look good in the output
summary(nhanes_complete, title = "Participant characteristics by missing status")

```




## Missingness patterns over variables



```{r, message =FALSE, warning =FALSE , echo=FALSE, eval=FALSE}

# This is a dendogram that shows which variables tend to be missing together:
  
s_nhanes <- a_nhanes[sample(1:nrow(a_nhanes), size=10000, repl=TRUE), ]
is_nhanes <- is.na(a_nhanes)*1

perc_miss <- round(apply(is_nhanes, 2, mean)*100,0)
colnames(is_nhanes)<-paste(colnames(is_nhanes),"(",perc_miss,")",sep="")
hobj <- hclust(dist(t(is_nhanes)))
plot(hobj, cex=0.5)
```



```{r, message =FALSE, warning =FALSE , echo=FALSE, fig.width==10}
aggr(a_nhanes[,c(demog_vars, life_vars, health_vars)],numbers=TRUE)
```

Missing values for each participant in the NHANES dataset is shown in the following figure, where the black lines correspond to observations with missing values.

```{r, message =FALSE, warning =FALSE , echo=FALSE}
a_nhanes %>%
  select(demog_vars, life_vars, health_vars, physio_vars) %>%
  naniar::vis_miss(sort_miss = TRUE, show_perc_col = TRUE) 
```


There are **`r n_var_miss(a_nhanes[,c(demog_vars, life_vars, physio_vars, health_vars)])`** independent variables with missing values in the dataset.

```{r, message =FALSE, warning =FALSE , echo=FALSE}
gg_miss_upset(a_nhanes[,c(demog_vars, life_vars, physio_vars, health_vars)], 
              nsets = 10,
              nintersects = NA)
```

Physiological variables (blood pressure, cholesterol) have the highest proportion of missingness. There does not seem to be a pattern of missingness across variables, other than the cholesterol (toal, HDL) variables.



In addition, we can explore missing data mechanisms and relationships between BMI and systolic blood pressure, included in the same scatterplot:


```{r, message =FALSE, warning =FALSE , echo=FALSE}
ggplot(a_nhanes,
       aes(x = bmi,
           y = sys)) +
 geom_miss_point() +
facet_wrap(~gender)
```

Missing values are seen across the same range of values for the other variable. There is an extreme value in BMI for males that is likely an entry error.


## (In)complete cases

This section presents patients with a least one missing value. First we list out patients with at least one missing value in a filterable table. 

```{r, message =FALSE, warning =FALSE , echo=FALSE}
cc <-
  a_nhanes %>%  
  select(-seqn,-mvpa) %>% 
  filter(!complete.cases(.))
DT::datatable(cc)
```

Then we report the pattern of missing for this set of patients.

```{r, message =FALSE, warning =FALSE , echo=FALSE}
cc %>%
  select(demog_vars, life_vars, health_vars, physio_vars) %>%
  naniar::vis_miss(sort_miss = TRUE, show_perc_col = TRUE) 


```




## Section session info

```{r, message =FALSE, warning =FALSE , echo=FALSE}
sessionInfo()
```