# Missing data


```{r, echo=FALSE, warning=FALSE, message=FALSE, echo=FALSE}
## Load libraries
library(here)
library(tidyverse)
library(Hmisc)
library(naniar) ## for missing data 
library(gt)
library(kableExtra)
library(DT)
library(arsenal)
library(VIM)


load(here::here("data", "a_nhanes.rda"))

#load(here::here("a_nhanes.rda"))



# Define sets of predictors 
vip_vars <-c("age", "gender")
demog_vars <- c("age", "gender", "educationadult")
life_vars <- c("bmi", "smokecigs" , "drinkstatus")
health_vars <- c("diabetes", "chf", "cancer" , "stroke")
physio_vars <- c("sys","lbxtc",  "lbdhdd")
pa_vars <- c( "tac" ,"tlac" ,"mvpa", "wt" , "tlac.1", "tlac.2","tlac.3","tlac.4","tlac.5","tlac.6","tlac.7","tlac.8","tlac.9","tlac.10","tlac.11","tlac.12" )

vars<-c(demog_vars,life_vars,health_vars, physio_vars, "tac" ,"tlac" ,"mvpa", "wt" )
```


## Per variable missingness

Number and percentage of missing. 

```{r, message =FALSE, warning =FALSE , echo=FALSE}
# devtools::install_github("rstudio/gt")
a_nhanes %>%
  select(-seqn) %>%
  miss_var_summary() %>%
  gt::gt() %>%
  gt::cols_label(
    variable = md("**Variable**"),
    n_miss = md("**Missing (count)**"),
    pct_miss = md("**Missing (%)**")
  ) %>%
  gt::fmt_number(
    columns = vars(pct_miss),
    decimals = 2
  )
```


Investigate for groups of variables:

```{r, message =FALSE, warning =FALSE , echo=FALSE}
a_nhanes %>%
  select() %>%
  mutate(any_variable_missing = ifelse(apply(a_nhanes,1,any_miss),NA,1)) %>%
  mutate(any_demographics_missing = ifelse(apply(a_nhanes[,demog_vars],1,any_miss),NA,1)) %>%
  mutate(any_vip_missing = ifelse(apply(a_nhanes[, vip_vars],1,any_miss),NA,1)) %>%
  mutate(any_lifestyle_missing = ifelse(apply(a_nhanes[,life_vars],1,any_miss),NA,1)) %>%
  mutate(any_health_missing = ifelse(apply(a_nhanes[,health_vars],1,any_miss),NA,1)) %>%
  mutate(any_physact_missing = ifelse(apply(a_nhanes[,pa_vars],1,any_miss),NA,1)) %>%
   mutate(any_lab_missing = ifelse(apply(a_nhanes[,physio_vars],1,any_miss),NA,1)) %>%
  miss_var_summary() %>%
  gt::gt() %>%
    gt::cols_label(
    variable = md("**Variable**"),
    n_miss = md("**Missing (count)**"),
    pct_miss = md("**Missing (%)**")
  ) %>%
  gt::fmt_number(
    columns = vars(pct_miss),
    decimals = 2
  )

```

Overall, 18% are missing when all variables are included in a model. There is no missingness in the VIPs and less than 1% in life style variables such as smoking and alcohol consumption.


```{r, message =FALSE, warning =FALSE , echo=FALSE}
a_nhanes %>%
  select(vars) %>%
  gg_miss_var(show_pct = TRUE)
```


```{r, results='asis', message =FALSE, warning =FALSE , echo=FALSE, eval=FALSE}

## Variable summaries for complete vs incomplete cases


a_nhanes$mvpamiss<-"incomplete"
idx<-which(is.na(a_nhanes$mvpa))
a_nhanes$mvpamiss[idx]<-"complete"


vars<-c(demog_vars, health_vars,  "tlac", "wt", "mvpamiss" )

nhanes_complete<-tableby(mvpamiss ~ ., data=a_nhanes[,vars], test=TRUE, numeric.stats=c("median","q1q3", "range"))

# need results='asis' above for this to look good in the output
summary(nhanes_complete, title = "Participant characteristics by missing status of MVPA")

```




## Missingness patterns over variables



```{r, message =FALSE, warning =FALSE , echo=FALSE, eval=FALSE}

# This is a dendogram that shows which variables tend to be missing together:
  
s_nhanes <- a_nhanes[sample(1:nrow(a_nhanes), size=10000, repl=TRUE), ]
is_nhanes <- is.na(a_nhanes)*1

perc_miss <- round(apply(is_nhanes, 2, mean)*100,0)
colnames(is_nhanes)<-paste(colnames(is_nhanes),"(",perc_miss,")",sep="")
hobj <- hclust(dist(t(is_nhanes)))
plot(hobj, cex=0.5)
```



```{r, message =FALSE, warning =FALSE , echo=FALSE}
aggr(a_nhanes[,c(demog_vars, life_vars, health_vars)],numbers=TRUE)
```


```{r, message =FALSE, warning =FALSE , echo=FALSE}
a_nhanes %>%
  select(demog_vars, life_vars, health_vars) %>%
  naniar::vis_miss(sort_miss = TRUE, show_perc_col = TRUE) 
```



## (In)complete cases

This section presents patients with a least one missing value. First we list out patients with at least one missing value in a filterable table. 

```{r, message =FALSE, warning =FALSE , echo=FALSE}
cc <-
  a_nhanes %>%  
  select(-seqn,-mvpa) %>% 
  filter(!complete.cases(.))
DT::datatable(cc)
```

Then we report the pattern of missing for this set of patients.

```{r, message =FALSE, warning =FALSE , echo=FALSE}
cc %>%
  select(demog_vars, life_vars, health_vars) %>%
  naniar::vis_miss(sort_miss = TRUE, show_perc_col = TRUE) 
```




## Section session info

```{r, message =FALSE, warning =FALSE , echo=FALSE}
sessionInfo()
```