# Univariate distributions

Univariate summary CRASH-2 dataset 

```{r, message =FALSE, warning =FALSE , echo=FALSE}
library(here)
library(tidyverse)
library(Hmisc)
source(here("R", "ida_plot_univar.R"))  ## function to plot univariate summaries. 

# Define sets of predictors by type - helpful for selection
demog_vars <- c("age", "sex")
pyhiso_vars <- c("sbp", "hr", "rr", "gcs", "cc")
injury_vars <- c("injurytime", "injurytype")

## Load the dataset.
load(here::here("data", "a_crash2.rda"))
```


## Data set overview

Using Hmisc describe function, provide an overview of the data set is provided including histograms of continuous variables.  

### Demographic variables

**TODO: Should we plot the marginal distribution of the outcome? **

```{r desc_a_crash2a, message = FALSE, results='asis', warning=FALSE, , echo=FALSE}
a_crash2 %>%
  dplyr::select(all_of(demog_vars)) %>%
  Hmisc::describe(descript = "Demographic variables") %>%
  Hmisc::html(size = 80)
```


### Physiological measurements

```{r desc_a_crash2b, message = FALSE, results='asis', warning=FALSE, , echo=FALSE}
a_crash2 %>%
  dplyr::select(all_of(pyhiso_vars)) %>%
  Hmisc::describe(descript = "Physiological measurements") %>%
  Hmisc::html(size = 80)
```

### Characteristics of injury

```{r desc_a_crash2c, message = FALSE, results='asis', warning=FALSE, echo=FALSE}
a_crash2 %>%
  dplyr::select(all_of(injury_vars)) %>%
  Hmisc::describe(descript = "Characteristics of injury") %>%
  Hmisc::html(size = 80)
```

## Categorical plots

A closer examination of the categorical predictors. 

```{r catplot, message=FALSE, warning =FALSE , echo=FALSE}
a_crash2 %>%
  dplyr::select(sex, injurytype) %>%
  dplyr::mutate_all(forcats::as_factor) %>%   
  dplyr::mutate_all(forcats::fct_explicit_na, "NA") %>%
  tidyr::pivot_longer(
    dplyr::everything(),
    names_to = "var",
    values_to = "value",
    values_drop_na = FALSE
  ) %>%
  dplyr::group_by(var, value) %>%
  dplyr::summarize(N = n()) %>%
  dplyr::mutate(
    freq = N / sum(N),
    pct = round((freq * 100), 1),
    axis_lab = paste0(value, ' ', '(', N, ', ', round(pct, digits = 1), '%)'),
    var_label = case_when(var == "sex" ~ "Sex",
                          var == "injurytype" ~ "Injury type")
  ) %>%
  ggplot(aes(
    x = reorder(axis_lab, pct),
    y = pct,
    label = pct
  )) +
  geom_point(color = "firebrick2",
             alpha = 0.6,
             size = 3) +
  ylab("Percentage (%)") +
  scale_y_continuous(limits = c(0, 100)) +
  labs(caption = "Number and percentage of subjects reported in brackets.\nNA = missing") +
  facet_wrap(~ var_label, ncol = 1, scales = "free_y") +
  coord_flip() +
  theme_minimal(base_size = 12) +
  theme(axis.title.y = element_blank(),
        panel.grid.minor = element_blank())

```

### Categorical ordinal plots

The Glasgow coma score, an ordinal categorical variable, is also displayed separately. 

```{r, message=FALSE, warning =FALSE , echo=FALSE}
a_crash2 %>%
  dplyr::select(gcs) %>%
  dplyr::mutate(gcs = as_factor(gcs)) %>%
  tidyr::pivot_longer(
    dplyr::everything(),
    names_to = "var",
    values_to = "value",
    values_drop_na = FALSE
  ) %>%
  dplyr::group_by(var, value) %>%
  dplyr::summarize(N = n()) %>%
  dplyr::mutate(
    freq = N / sum(N),
    pct = round((freq * 100), 1),
    value = as_factor(value),
    axis_lab = as_factor(paste0(value, ' ', '(', N, ', ', round(pct, digits = 1), '%)'))
  ) %>%
  ggplot(aes(
    x = axis_lab,
    y = pct,
    label = pct
  )) +
  geom_point(color = "firebrick2",
             alpha = 0.6,
             size = 3) +
  ylab("Percentage (%)") +
  scale_y_continuous(limits = c(0, 100)) +
  labs(title = "Glasgow coma score (point scale)",
    caption = "Number and percentage of subjects reported in brackets.\nNA = missing") +
  coord_flip() +
  theme_minimal(base_size = 12) +
  theme(axis.title.y = element_blank(),
        panel.grid.minor = element_blank())
```



## Continuous plots 

A closer examination of continuous predictors. 

```{r, message =FALSE, warning =FALSE , echo=FALSE, fig.width=9}

a_crash2 %>%
  dplyr::select(age, sbp, rr, cc, hr, injurytime) %>%
  dplyr::mutate() %>%
  dplyr::mutate_all(as.numeric) %>%
  tidyr::pivot_longer(dplyr::everything(),
                      names_to = "var",
                      values_to = "value") %>%
  dplyr::mutate(
    var_label = case_when(
      var == "age" ~ "Age (years)",
      var == "sbp" ~ "Systolic blood pressure (mmHg)",
      var == "hr" ~  "Heart rate (1/min)",
      var ==  "rr" ~ "Respiratory rate (1/min)",
      var ==  "cc" ~ "Central capillary refill time (seconds)", 
      var == "injurytime" ~ "Injury time (days)"
    )) %>%
  ggplot(aes(value)) +
  ylab("Number of subjects") +
  geom_histogram(bins = 200, alpha = 0.6, fill = "firebrick2") +
  facet_wrap(~ var_label, scales = "free", ncol = 2) +
  theme_minimal() +
  theme(axis.title.x = element_blank(),
        panel.grid.minor = element_blank())

```


There is evidence of digit preference. Explore further with targeted summaries. 

More detailed univariate summaries for the variables of interest are also provided below. 

### Age

```{r, fig.width=8, fig.height=3.5, fig.cap= "Distribution of subject age [years]" , echo=FALSE}
## This is a function the plots a strip plot, histogram and boxplot, including five number summary. 
ida_plot_univar(a_crash2, "age")
```


### Blood pressure

```{r, fig.width=8, fig.height=3.5, fig.cap= "Distribution of SBP" , echo=FALSE}
ida_plot_univar(a_crash2, "sbp")
```


### Respiratory rate


```{r, fig.width=8, fig.height=3.5, fig.cap= "Distribution of respiratory rate" , echo=FALSE}
ida_plot_univar(a_crash2, "rr", n_dodge = 2)
```


### Heart rate


```{r, fig.width=8, fig.height=3.5, fig.cap= "Distribution of heart rate", echo=FALSE}
ida_plot_univar(a_crash2, "hr")
```


### Central capillary refill time


```{r, fig.width=8, fig.height=3.5, fig.cap= "Distribution of Central capillary refill time", echo=FALSE}
ida_plot_univar(a_crash2, "cc")
```


### Hours since injury


```{r, fig.width=8, fig.height=3.5, fig.cap= "Distribution of hours since injury", echo=FALSE}
ida_plot_univar(a_crash2, "injurytime", n_dodge = 3)
```



## Session info

```{r}
sessionInfo()
```


